\documentclass[11pt]{article}

% ── packages ────────────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm, bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{mdframed}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{enumitem}

% ── todo macro ──────────────────────────────────────────────────────────────
\newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO:} #1}}

% ── theorem environments ────────────────────────────────────────────────────
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}

% ── auto-generated macro files ──────────────────────────────────────────────
\input{values}       % \srankSeventyM, \bonusRKfive..., \bitfitFinalLoss, ...
% lean_status is \input-ted in the body (§3) because it also typesets a table

% ── title ───────────────────────────────────────────────────────────────────
\title{Axiomatic Bounds on LoRA Alignment Geometry:\\
A Task-Intrinsic Dimensional Floor Across Pythia 70M–1B}
\author{Debanjan Basu}
\affil{Independent Research}
\date{\today}

\begin{document}
\maketitle

% ── abstract ────────────────────────────────────────────────────────────────
\begin{abstract}
The phenomenological success of Low-Rank Adaptation (LoRA) and Direct
Preference Optimization (DPO) at immense parameter scales has outpaced
their theoretical foundations. Current practice treats the adapter rank
$r$ and scaling factor $\alpha$ as hyperparameters that dictate the
expressivity of the alignment update. In this work, we present an
axiomatic framework and empirical scaling study mathematically decoupling
update \emph{energy} from alignment \emph{geometry}. Using Lean 4 and
\texttt{Mathlib}, we formally verify that while the choice of
$\alpha$-scaling strictly bounds the Frobenius energy of the update---
formally proving the $\alpha/\sqrt{r}$ RsLoRA scheme as a guaranteed
norm-conservation mechanism---neither hyperparameter alters the update's
intrinsic stable rank ($srank$). Empirically, we sweep the Pythia suite
from 70M to 1B parameters and find a \emph{task-intrinsic dimensional
constant} of $srank \approx \srankFitTaskConst \pm 0.5$, invariant across
three orders of magnitude of model width (RMS residual
$\srankFitTaskRMS$, vs.\ $\srankFitSqrtRMS$ for a $1/\sqrt{d}$ acoustic
scaling fit). LoRA updates concentrate in a 3--4 dimensional $\gamma$-rudder
that overlaps the base weight's top right-singular subspace at
$\bonusRKsrankFourTenMDPO\times$ random chance (Pythia-410M, DPO). We
isolate a distinct objective-dependent signal, finding that CLM adapters
exhibit higher top-$k$ subspace overlap than DPO adapters at $k{=}5$
($\bonusRKfiveOneBCLMavgSeeds\times$ vs.\
$\bonusRKfiveOneBDPOavgSeeds\times$ at 1B, averaged across seeds 42
and 117), quantifying the geometric cost of
contrastive preference routing. Through high-resolution ablation of
bias-only fine-tuning (BitFit), we demonstrate that only
$\bitfitLearningRatio$ of the DPO loss reduction is gauge-accessible
via additive biases; the remaining majority strictly requires novel
low-rank geometric matrix modification. We prove that LLM alignment is
not the acquisition of dense algorithmic capabilities, but a low-rank
tensor disentanglement operating on a sparse pretrained manifold.
\end{abstract}

% ── §1 Introduction ─────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

Low-Rank Adaptation (LoRA) \citep{hu2021lora} and Direct Preference
Optimization (DPO) \citep{rafailov2023dpo} have become the de-facto
compute-efficient alignment stack for large language models. In practice,
practitioners tune the adapter rank $r$, the scaling factor $\alpha$,
and the learning rate largely by folklore: $r=64$ or $r=128$ is
``enough'', $\alpha = 2r$ is a common default, and RsLoRA's
$\alpha/\sqrt{r}$ variant \citep{kalajdzievski2023rsLoRA} is adopted on
the basis of empirical speed-up rather than theoretical clarity. Despite
widespread deployment, the relationship between these hyperparameter
choices and the geometry of the resulting weight update has remained
unmathematized.

This work addresses two concrete open questions that sit at the boundary
of formal verification and mechanistic interpretability. First, does the
choice between $\alpha/r$ and $\alpha/\sqrt{r}$ scaling genuinely alter
the shape of the alignment update, or does it modulate only its
magnitude? Answering this question rigorously requires distinguishing the
\emph{energy} (Frobenius norm squared) of $\Delta W$ from its
\emph{geometry} (intrinsic dimensionality, as measured by stable rank).
Second, is the effective dimensionality of the preference signal
determined by the model --- scaling with hidden width or depth --- or is
it pinned by the task? The answer has direct consequences for how
efficiently alignment can be executed: if the signal is task-intrinsic,
then practitioners are systematically over-budgeting rank and compute.

We make the following contributions.

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Lean 4 formalization of three theorems} (Table~\ref{tab:lean-status}).
    We mechanically verify in Lean 4 with \texttt{Mathlib} that
    standard LoRA's energy bound decays as $1/r$ whereas RsLoRA's
    remains rank-invariant (Theorem~\ref{thm:energy-bounds}); that the
    stable rank of any matrix is invariant under scalar multiplication
    and therefore cannot be altered by any choice of $\alpha$
    (Theorem~\ref{thm:geom-invariance}); and that the accumulated SGD
    weight update over $N$ steps has matrix rank at most $N$
    (Theorem~\ref{thm:outer-rank}).

  \item \textbf{Empirical sweep of the $\gamma$-rudder across the Pythia suite.}
    We measure the subspace overlap $\text{bonus}_R(k)$ between
    $\Delta W$'s top-$k$ right-singular vectors and the base weight's
    top-$k$ right-singular vectors, sweeping from 70M to 1B parameters.
    Across this three-order-of-magnitude range, the stable rank of LoRA
    adapters trained with DPO clusters tightly around a
    \emph{task-intrinsic dimensional constant} of
    $srank \approx \srankFitTaskConst \pm 0.5$, invariant to model
    width (Table~\ref{tab:gamma-scaling}, Table~\ref{tab:scaling-fit}).

  \item \textbf{Gauge-theory falsification via BitFit-DPO ablation.}
    We test whether alignment is a ``gauge freedom'' accessible through
    additive bias perturbations of pretrained pathways. Training only
    bias parameters of Pythia-410M under DPO achieves a learning ratio
    of just $\bitfitLearningRatio$ relative to LoRA; extending training
    nearly doubles the step budget without producing a phase transition
    (Table~\ref{tab:bitfit}, Table~\ref{tab:autopsy}).
\end{enumerate}

The paper is organized as follows. Section~\ref{sec:prelim} establishes
notation and background. Section~\ref{sec:theory} states and proves the
three theorems, with references to the Lean artifact.
Section~\ref{sec:empirical} presents the Pythia sweep, scaling-fit
analysis, and BitFit ablation. Section~\ref{sec:discussion} connects the
findings to the broader literature on intrinsic dimensionality, PiSSA
initialization, and the safety implications of alignment geometry.
Section~\ref{sec:limitations} acknowledges scope limitations and
identifies directions for follow-on work.


% ── §2 Preliminaries ────────────────────────────────────────────────────────
\section{Preliminaries}
\label{sec:prelim}

\subsection{LoRA and RsLoRA}

Let $W \in \mathbb{R}^{m \times n}$ be a weight matrix of a pretrained
transformer. Low-Rank Adaptation \citep{hu2021lora} augments $W$ with a
trainable update $\Delta W = BA$, where $B \in \mathbb{R}^{m \times r}$
and $A \in \mathbb{R}^{r \times n}$ with $r \ll \min(m, n)$. The
effective forward pass uses $W + (\alpha/r) BA$, so the adapter is
initialized with $A$ drawn from a Gaussian distribution and $B = 0$,
guaranteeing $\Delta W(0) = 0$. The scalar $\alpha/r$ controls the
magnitude of the update at initialization.

RsLoRA \citep{kalajdzievski2023rsLoRA} replaces this scaling with
$\alpha / \sqrt{r}$, arguing that the standard $\alpha/r$ schedule
systematically reduces signal energy as rank grows, while
$\alpha/\sqrt{r}$ maintains constant expected-norm behavior. We formalize
and verify this distinction in Theorem~\ref{thm:energy-bounds}.

\subsection{DPO and the CLM Control}

Direct Preference Optimization \citep{rafailov2023dpo} fine-tunes a
language model $\pi_\theta$ to prefer accepted completions $y_w$ over
rejected completions $y_l$ given a prompt $x$, without explicit reward
modeling. The training objective is:
\begin{equation}
\mathcal{L}_{\text{DPO}}(\theta) = -\mathbb{E}_{(x, y_w, y_l)}\left[
  \log \sigma\!\left( \beta \log \frac{\pi_\theta(y_w \mid x)}{\pi_{\text{ref}}(y_w \mid x)}
    - \beta \log \frac{\pi_\theta(y_l \mid x)}{\pi_{\text{ref}}(y_l \mid x)}
  \right)
\right]
\end{equation}
where $\pi_{\text{ref}}$ is a frozen copy of the pretrained model and
$\beta = 0.1$ controls the strength of the KL penalty. We train on the
\texttt{Anthropic/hh-rlhf} dataset \citep{bai2022hhrlhf} for 800 steps,
using the ``chosen'' field of each example as $y_w$ and the ``rejected''
field as $y_l$.

To isolate the geometric footprint of contrastive preference optimization
from the footprint of domain adaptation, we train a \emph{CLM control}:
standard next-token-prediction on the same ``chosen'' completions, with
identical LoRA configuration, learning rate, and step count. Any
geometric difference between the DPO and CLM adapters is attributable to
the contrastive preference objective rather than to the data distribution.

\subsection{Stable Rank}
\label{sec:prelim-srank}

For a matrix $M \in \mathbb{R}^{m \times n}$, the \emph{stable rank} is
defined as:
\begin{equation}
srank(M) \;=\; \frac{\|M\|_F^2}{\|M\|_2^2}
\end{equation}
where $\|M\|_F$ is the Frobenius norm and $\|M\|_2$ is the spectral
(largest singular value) norm. Stable rank is a continuous relaxation of
matrix rank, satisfying $srank(M) \leq \mathrm{rank}(M)$, and measures
the effective number of dimensions over which the matrix concentrates its
energy. We use stable rank as the primary geometric observable throughout
this paper because it is scale-invariant (Theorem~\ref{thm:geom-invariance}),
interpretable, and sensitive to the anisotropy of the singular value
spectrum.

\subsection{The $\gamma$-Metric (Subspace Overlap)}

Let $V_W \in \mathbb{R}^{n \times k}$ be the matrix of top-$k$
right-singular vectors of $W$, and let $V_{\Delta W} \in
\mathbb{R}^{n \times k}$ be the analogous matrix for $\Delta W$. We
define the right-side subspace overlap bonus at rank $k$ as:
\begin{equation}
\text{bonus}_R(k) \;=\; \frac{\|V_W^\top V_{\Delta W}\|_F^2 / k}{k / d_{\text{in}}}
\;=\; \frac{d_{\text{in}} \,\|V_W^\top V_{\Delta W}\|_F^2}{k^2}
\end{equation}
where $d_{\text{in}} = n$ is the input dimension of the weight matrix.
The denominator $k / d_{\text{in}}$ is the expected overlap fraction for
two independently drawn uniformly random $k$-dimensional subspaces of
$\mathbb{R}^{d_{\text{in}}}$. A value of $\text{bonus}_R(k) = 1$
indicates no alignment beyond chance; values above 1 indicate that the
LoRA adapter preferentially modifies the directions that $W$ already
operates on most strongly. An analogous left-side metric
$\text{bonus}_L(k)$ uses the left-singular vectors $U_W$ and
$U_{\Delta W}$ with denominator $k / d_{\text{out}}$. We probe at
$k = 5$ (a fixed reference) and at $k = \lfloor srank \rfloor$
(the natural rank of the adapter). Throughout the paper we refer to the
cluster of dominant adapter directions as the \emph{$\gamma$-rudder}.

\subsection{The Pythia Model Suite}

We conduct all scaling experiments on the Pythia family
\citep{biderman2023pythia}, which provides a matched set of GPT-NeoX
architecture models trained on the Pile with identical hyperparameters
across scales. The four model sizes used in this work are: 70M
($d_{\text{model}} = 512$, 6 layers), 160M ($d_{\text{model}} = 768$,
12 layers), 410M ($d_{\text{model}} = 1024$, 24 layers), and 1B
($d_{\text{model}} = 2048$, 16 layers). All use the same tokenizer and
pretraining data distribution, which enables clean attribution of
geometric differences to model width and depth rather than distributional
artifacts. All LoRA adapters in this paper target the
\texttt{attention.query\_key\_value} module at every transformer layer.


% ── §3 Theoretical Framework ────────────────────────────────────────────────
\section{Theoretical Framework}
\label{sec:theory}

% lean_status defines \thmStat... macros used below AND typesets the status table
\input{lean_status}

\subsection{Stable Rank and the Alignment Geometry}

To distinguish between the magnitude of a parameter update and its
dimensional structure, we use the \emph{stable rank}:
\begin{equation}
    srank(M) \;=\; \frac{\|M\|_F^2}{\|M\|_2^2}
\end{equation}
where $\|M\|_F$ is the Frobenius norm and $\|M\|_2$ is the spectral
(operator) norm. Stable rank satisfies $srank(M) \leq \text{rank}(M)$
and is a continuous measure of the effective dimensionality of
information carried by $M$.

\subsection{Decoupling Energy and Geometry}

Current heuristic practices in LLM alignment treat the choice of rank
$r$ and scaling factor $\alpha$ as modulators of both scale and
expressivity. Let $B \in \mathbb{R}^{m \times r}$ and
$A \in \mathbb{R}^{r \times n}$ be the LoRA adapter factors. Standard
LoRA defines the update as $\Delta W_{\mathrm{LoRA}} = \frac{\alpha}{r} BA$,
whereas RsLoRA \citep{kalajdzievski2023rsLoRA} defines it as
$\Delta W_{\mathrm{RsLoRA}} = \frac{\alpha}{\sqrt{r}} BA$. Using the
Lean~4 interactive theorem prover and the \texttt{Mathlib} library, we
formalize the following bounds:

\begin{theorem}[Update Energy Bounds] \label{thm:energy-bounds}
Under the boundedness assumption $\|BA\|_F^2 \leq c\,r$ (which holds
deterministically for any bounded factor norms and with high probability
under standard Gaussian initialization):
\begin{enumerate}[label=(\alph*)]
\item \textbf{Standard LoRA decay:}
  $\|\Delta W_{\mathrm{LoRA}}\|_F^2 \leq \alpha^2 \frac{c}{r}$.
  The upper bound on signal energy decays proportionally to $1/r$.
\item \textbf{RsLoRA conservation:}
  $\|\Delta W_{\mathrm{RsLoRA}}\|_F^2 \leq \alpha^2 c$.
  The signal energy remains strictly bounded and invariant to $r$.
\end{enumerate}
(Proof mechanized in Lean~4: \texttt{loraUpdate\_frob\_decays}
\thmStatOfLoraUpdateFrobDecays~and \texttt{rsLoraUpdate\_frob\_bounded}
\thmStatOfRsLoraUpdateFrobBounded.)
\end{theorem}

\begin{mdframed}[backgroundcolor=gray!10, roundcorner=5pt, innertopmargin=10pt, innerbottommargin=10pt]
\begin{theorem}[Geometric Invariance under Scalar Transformation] \label{thm:geom-invariance}
For any non-zero scalar $c \in \mathbb{R} \setminus \{0\}$ and any
matrix $M$:
\begin{equation}
    srank(c \cdot M) = srank(M)
\end{equation}
\begin{proof}
Mechanized in Lean~4 (\texttt{stableRank\_smul\_invariant}
\thmStatOfStableRankSmulInvariant). The Frobenius norm squared scales
quadratically by direct summation:
$\|cM\|_F^2 = \sum_{i,j}(cM_{ij})^2 = c^2 \|M\|_F^2$
(\texttt{frobeniusSq\_smul} \thmStatOfFrobeniusSqSmul). The spectral
norm squared scales identically via the continuous-linear-map operator
norm, composing \texttt{Matrix.toEuclideanLin} with
\texttt{LinearMap.toContinuousLinearMap}, then applying absolute
homogeneity and squaring (\texttt{spectralSq\_smul}
\thmStatOfSpectralSqSmul). The ratio therefore reduces to
$(c^2 \|M\|_F^2) / (c^2 \|M\|_2^2) = srank(M)$ for $c \neq 0$.
\end{proof}
\end{theorem}
\end{mdframed}

\begin{corollary}[Hyperparameter Geometric Independence] \label{cor:hyp-indep}
$srank(\Delta W_{\mathrm{LoRA}}) = srank(\Delta W_{\mathrm{RsLoRA}}) = srank(BA)$.
(See §\ref{sec:empirical} for empirical validation across Pythia-410M
and 1B with two random seeds.)
\end{corollary}

\textbf{Implication.} Theorem~\ref{thm:geom-invariance} rigorously
decouples scale from dimension. Tuning $\alpha$ or choosing between
standard LoRA and RsLoRA strictly modulates the magnitude of the
alignment routing without altering its intrinsic dimensional structure.

\subsection{Rank Bounds on Accumulated Updates}

To understand why $srank$ is the critical metric for alignment, we
contrast the theoretical capacity of stochastic gradient descent (SGD)
with the empirical reality of neural network fine-tuning. In a single
backpropagation step, the weight update is proportional to the tensor
outer product of the error gradient $\delta$ and the input activation
$x$: $\Delta W \propto \delta \otimes x^\top$.

\begin{theorem}[Outer-Product Rank Bounds] \label{thm:outer-rank}
Let $\Delta W = \sum_{i=1}^{N} \delta_i \otimes x_i^\top$ be the
accumulated weight update over $N$ training examples.
\begin{enumerate}[label=(\alph*)]
\item \textbf{Single update:} $\mathrm{rank}(\delta \otimes x^\top) \leq 1$.
\item \textbf{Accumulated update:}
  $\mathrm{rank}(\sum_{i=1}^{N} \delta_i \otimes x_i^\top) \leq N$.
\end{enumerate}
(Proof mechanized in Lean~4 via matrix factorization and subadditivity
of rank: \texttt{rank\_outer\_product\_le\_one}
\thmStatOfRankOuterProductLeOne~and \texttt{rank\_sum\_outer\_products\_le}
\thmStatOfRankSumOuterProductsLe.)
\end{theorem}

For full-parameter fine-tuning over a standard DPO run of 800 steps at
a batch size of 16, Theorem~\ref{thm:outer-rank} yields an algebraic
upper bound of approximately $\mathbf{\sgdBound}$ on $\mathrm{rank}(\Delta W)$.
Low-rank adaptation (LoRA) imposes an additional structural cap by
construction, since $\Delta W_{\mathrm{LoRA}} = BA$ factors through a
bottleneck of dimension $r$: $\mathrm{rank}(\Delta W_{\mathrm{LoRA}}) \leq r$.
For our configured $r = \mathbf{\loraCap}$, this alone represents a
$\mathbf{{\sim}\sgdToLoRAcompression\times}$ compression below the SGD
accumulation bound.

However, as demonstrated in §\ref{sec:empirical}, the empirical stable
rank collapses to a narrow bandwidth of $srank \approx \mathbf{3\text{--}5}$---
a further $\mathbf{{\sim}\loraToSrankCompression\times}$ reduction below
LoRA's hard cap, and a combined
$\mathbf{{\sim}\totalCompression\times}$ reduction below the
full-parameter SGD accumulation bound. This disparity provides formal
evidence for tensor disentanglement: alignment operates on a highly
sparse pretrained manifold where thousands of updates destructively
interfere, isolating a low-dimensional $\gamma$-rudder that routes
preference logic.

\paragraph{Lean~4 artifact statement.} All theorems in this section are
formally verified using the Lean~4 interactive theorem prover and
\texttt{Mathlib}. The source file \texttt{SubspaceOverlap.lean} contains
\leanProvenCount~real-proven theorems with zero \texttt{sorry} in the
scaling branch; \leanSorryCount~targeted \texttt{sorry} stubs mark
literature-terminology placeholders for downstream work
(Table~\ref{tab:lean-status}).


% ── §4 Empirical Ablations ──────────────────────────────────────────────────
\section{Empirical Ablations}
\label{sec:empirical}

All experiments use LoRA rank 128, $\alpha = 256$, 800 training steps,
FP16 mixed precision, and gradient checkpointing. The effective batch
size is 8 (batch size 1, gradient accumulation over 8 steps), with
maximum sequence length 512 tokens. DPO training uses $\beta = 0.1$
with a frozen reference model equal to the pretrained checkpoint. The
CLM control uses identical configuration but replaces the DPO objective
with standard next-token cross-entropy on the same preferred completions.
All experiments run on a single NVIDIA RTX 3060 12~GB GPU. Adapters are
applied to the \texttt{attention.query\_key\_value} projection at every
layer of each model.

\input{tables}

\subsection{The $\gamma$-Rudder Across Scales}

Table~\ref{tab:gamma-scaling} reports stable rank and subspace-overlap
bonus factors for every model and training objective in the sweep. The
central observation is that the DPO stable rank remains within a
narrow band across the full 70M-to-1B range: $\srankSeventyM$ at 70M,
$\srankOneSixtyM$ at 160M, $\srankFourTenM$ at 410M, and
$\srankOneBDPOsFortyTwo$ (seed 42) / $\srankOneBDPOsOneSeventeen$
(seed 117) at 1B. Crucially, this ordering is non-monotone: the 410M
model has a higher srank than the smaller 160M model, and the 1B model
sits below both. This pattern directly refutes a smooth width-scaling
hypothesis; no $c/d^\gamma$ curve with $\gamma > 0$ fits a non-monotone
sequence without overfitting to noise.

The right-side subspace overlap at the natural rank,
$\text{bonus}_R(srank)$, forms a universal band across the sweep: from
$\bonusRKsrankSeventyM\times$ random at 70M, peaking at
$\bonusRKsrankOneSixtyM\times$ at 160M, and remaining elevated at
$\bonusRKsrankFourTenMDPO\times$ at 410M and
$\bonusRKsrankOneBDPOsFortyTwo\times$ at 1B. That is, every model in
the suite concentrates its alignment update in the same right-singular
subspace that the base weight already uses most actively --- at least
$5\times$ above chance across all scales. This is the $\gamma$-rudder
signature.

Reproducibility of the 1B results is confirmed by a second seed: the
DPO srank varies by less than $2\%$ across seeds 42 and 117, and the
$\text{bonus}_R(5)$ varies by less than $8\%$. The CLM control at 1B
exhibits a consistently higher $\text{bonus}_R$ than DPO at $k=5$
($\bonusRKfiveOneBCLMavgSeeds\times$ vs.\ $\bonusRKfiveOneBDPOavgSeeds\times$
on average across seeds), quantifying the geometric cost of contrastive
preference routing: DPO must partially exit the pretrained top-$k$ input
subspace to encode the contrast between preferred and rejected
completions. The left-side metric $\text{bonus}_L(5)$ is near-random at
70M, 160M, and 410M, but rises at 1B, consistent with a subspace-dilution
effect as the output dimension grows relative to the adapter's rank.

\subsection{Task-Intrinsic Scaling Fit}

Table~\ref{tab:scaling-fit} formalizes the scaling comparison by fitting
three candidate forms to the four measured DPO stable-rank values. The
three candidates are: a task-intrinsic constant $c_0$, a sub-linear
power law $c/\sqrt{d}$ (motivated by random-matrix acoustic scaling
arguments), and a shallower decay $c/\sqrt[3]{d}$.

The task-intrinsic constant achieves an RMS residual of
$\srankFitTaskRMS$ stable-rank units, substantially lower than both
the $1/\sqrt{d}$ fit ($\srankFitSqrtRMS$) and the $1/\sqrt[3]{d}$ fit
($\srankFitCbrtRMS$). The best-fit constant is $\srankFitTaskConst$.
This is not merely a consequence of the small number of data points: the
non-monotone ordering of the measured values (70M~$= \srankSeventyM$,
160M~$= \srankOneSixtyM$, 410M~$= \srankFourTenM$,
1B~$= \srankOneBDPOsFortyTwo$) provides positive evidence
that the data are drawn from a distribution with a task-set floor and
architecture-level variance of approximately $\pm 0.5$ around it, rather
than from a smooth monotone function of $d$. A monotone decreasing fit
would need to explain why the 410M model exceeds the 70M model --- which
it cannot do without attributing the observation to sampling noise.

The implication is that alignment dimensionality is a property of the
preference task and dataset, not of the model architecture. For the
\texttt{hh-rlhf} task family studied here, the signal requires
approximately $\srankFitTaskConst$ effective dimensions per QKV adapter,
independent of model scale. This has direct practical consequence: LoRA
ranks above 8 represent significant over-provisioning for preference
learning at this data scale.

\subsection{BitFit-DPO: Gauge-Theory Falsification}

The $\gamma$-rudder findings raise a falsifiability question: could the
observed subspace alignment be an artifact of the pretrained model
already encoding alignment-relevant directions as additive biases,
accessible without new weight-matrix geometry? We test this with a
bias-only DPO run on Pythia-410M following the BitFit protocol
\citep{benzaken2021bitfit}: all weight matrices are frozen, and only the
$\bitfitTrainableParams$ bias parameters ($\bitfitTrainableFrac\%$ of
total model parameters) are trained under the DPO objective for 800
steps on \texttt{hh-rlhf}.

Table~\ref{tab:bitfit} reports the result. BitFit achieves a final loss
of $\bitfitFinalLoss$ (last-10-step mean), compared with the LoRA-DPO
endpoint of $\bitfitLoRAReference$. The learning ratio --- the fraction
of LoRA's total loss reduction captured by BitFit --- is
$\bitfitLearningRatio$. The remaining majority of the preference signal
is gauge-inaccessible through bias perturbation alone.

To test for the possibility of a punctuated phase transition (grokking),
we extended the BitFit run to $\bitfitExtLastStep$ steps, nearly doubling
the training budget. The extended run reaches a final loss of
$\bitfitExtFinalLoss$ with no discontinuous improvement visible in the
training curve, confirming saturation rather than delayed convergence.
Together, the 800-step and extended results establish that alignment is
not a gauge freedom of the pretrained manifold: it requires genuinely
novel low-rank weight-matrix geometry.

\subsection{Bias-Theory Autopsy: Is $\gamma$ a Hidden Gain?}

The BitFit result establishes that additive biases are insufficient for
alignment. A subtler residual hypothesis is that the LoRA adapter,
despite being expressed as a low-rank matrix update, might be
approximating a diagonal rescaling of the base weight --- for example,
the per-channel gain modulation performed by LayerNorm's learnable
$\gamma$ parameter. Under this hypothesis, the apparent geometric
novelty of the $\gamma$-rudder would be illusory: the adapter would
simply be re-weighting the base weight's own columns or rows.

We test this with a bias-theory autopsy on the Pythia-410M adapters.
For each trained $\Delta W$ on \texttt{attention.query\_key\_value}, we
solve for the best-fit diagonal gains in two senses: a
right-multiplicative fit $\Delta W \approx W \,\mathrm{diag}(g_R)$,
which corresponds to per-input-channel rescaling (the LayerNorm-$\gamma$
analog), and a left-multiplicative fit $\Delta W \approx
\mathrm{diag}(g_L)\, W$, which corresponds to per-output-channel gain.
The Frobenius-squared residual fraction measures how much of $\Delta W$
lies outside the best diagonal approximation.

Table~\ref{tab:autopsy} reports the results across three adapter
variants (DPO $r=16$, DPO $r=128$, CLM $r=128$). The right-mult
residual fraction is at least $\autopsyResidualMinPct\%$ in every run;
the left-mult residual is similarly extreme. Less than $0.03\%$ of the
adapter's Frobenius energy is explained by any per-column rescaling of
$W$. Combined with the BitFit-DPO failure, this definitively rules out
the gauge-transform hypothesis: the $\gamma$-rudder is genuinely novel
geometric rewiring of the weight matrix, not a hidden bias or gain
modulation of existing pretrained directions.


% ── §5 Discussion ───────────────────────────────────────────────────────────
\section{Discussion}
\label{sec:discussion}

\paragraph{Connection to PiSSA.}
PiSSA \citep{meng2024pissa} initializes the LoRA factors by decomposing
$W$ via SVD and using the top-$r$ singular components as the initial
$B$ and $A$ matrices, leaving the residual low-energy components frozen.
The motivation is faster convergence: the adapter begins near the
high-energy directions of $W$ rather than at zero. Our $\gamma$-rudder
findings provide a mechanistic explanation for why this initialization
accelerates training: SGD with the standard (zero) initialization
converges toward the same destination --- as quantified by
$\text{bonus}_R(srank) \geq 5\times$ random across all scales --- but
takes more steps to arrive. The gradient signal has an implicit attractor
in the pretrained top-$k$ right-singular subspace, and PiSSA simply
places the adapter at that attractor from step zero. This observation
suggests that subspace-alignment is a universal feature of LoRA
optimization on language tasks, not an artifact of PiSSA's design.

\paragraph{Connection to intrinsic dimensionality.}
\citet{aghajanyan2020intrinsic} showed that the intrinsic dimensionality
of fine-tuning objectives for NLP tasks is orders of magnitude smaller
than the parameter count --- as few as a few hundred trainable parameters
suffice to recover most task performance. Our task-intrinsic floor of
$srank \approx \srankFitTaskConst$ sharpens this finding for preference
learning at the per-module level: each QKV adapter effectively uses
approximately $\srankFitTaskConst$ dimensions, regardless of the
adapter's nominal rank. The gap between the configured rank
($r = \loraCap$) and the observed srank reflects massive destructive
interference among outer-product updates, consistent with the
${\sim}\loraToSrankCompression\times$ compression ratio noted in
Section~\ref{sec:theory}.

\paragraph{Connection to sparse autoencoders and monosemanticity.}
The observation that the 160M model achieves
$\text{bonus}_R(srank) = \bonusRKsrankOneSixtyM\times$ --- the highest
in the sweep --- is intriguing in light of work on feature
disentanglement in mid-scale models \citep{bricken2023monosemanticity}.
Mid-scale models may represent a regime where the base weight's singular
structure is unusually well-aligned with the directions required for
preference signal, possibly because their limited capacity forces them to
represent features more monosemantically during pretraining. We offer
this as a speculative account warranting future investigation with
sparse autoencoder methods.

\paragraph{Safety implications.}
The BitFit-DPO failure provides a formal argument against the
``gauge-transform catastrophe'' interpretation of alignment
reversibility: if alignment were a simple bias shift of pretrained
pathways, a bias-only perturbation of equivalent parameter count would
suffice to undo it. Our result establishes that the alignment signal
is not gauge-accessible --- it is encoded as specific low-rank weight
geometry. Reversing alignment therefore requires undoing the exact
geometric rewiring applied during DPO, which is a more complex and
detectable operation than a bias flip. This supports a cautiously
optimistic view of alignment robustness against simple additive
adversarial perturbations.

\paragraph{Compute-efficient RLHF.}
The empirical srank floor of $\srankFitTaskConst \pm 0.5$ across
the Pythia suite implies that, for \texttt{hh-rlhf}-scale preference
tasks, LoRA ranks $r \geq 8$ are already in the regime of significant
over-provisioning: the gradient concentrates into approximately
$\srankFitTaskConst$ effective dimensions regardless. Current practice
of $r = 64$ or $r = 128$ wastes compute on adapter dimensions that
contribute negligible energy to the final $\Delta W$. Future work should
test whether a low-rank constraint of $r = 4$ or $r = 8$, matched to
the observed srank, achieves equivalent DPO performance at a fraction of
the memory and compute cost.


% ── §6 Limitations ──────────────────────────────────────────────────────────
\section{Limitations and Future Work}
\label{sec:limitations}

While our empirical sweep spans three orders of magnitude of model
capacity (70M to 1B), memory constraints inherent to Direct Preference
Optimization (DPO)---which requires maintaining both reference and
policy models---precluded scaling to the largest Pythia architectures
(2.8B, 6.9B, 12B) under strict FP16 experimental controls. Future work
must investigate whether the observed task-intrinsic dimensional floor
($srank \approx \srankFitTaskConst$) holds absolutely at the 10B+ scale,
or if extreme manifold sparsity induces further geometric collapse.
Integrating QLoRA (4-bit quantization) was outside the scope of this
work to keep the spectral measurements of the $\gamma$-rudder
unambiguous; characterizing the geometric degradation of quantized
alignment remains a critical open problem.

Further limitations: (i) all runs use Pythia; cross-architecture
replication (OLMo, Qwen, Llama) is queued but not reported here;
(ii) we study DPO and next-token-prediction only; other preference
objectives (IPO, KTO, ORPO) may induce different $\gamma$-rudder
geometries; (iii) the $\gamma$ metric is applied to
\texttt{attention.query\_key\_value} modules only; cross-module
comparison (\texttt{dense}, MLP) is deferred.


% ── bibliography ────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\bibliography{references}


\end{document}
